---
title: "RSA sandbox"
subtitle: "snippets and toy models (code here not tested extensively)"
author: "timothy leffel // `tjleffel@gmail.com` // `http://lefft.xyz`"
date: "updated nov03/2017"
output: html_document
---

### setup 

```{r setup}
lefftpack::lazy_setup()
# or use: 
# library("magrittr"); library("dplyr"); library("reshape2"); library("ggplot2")
```



### tweaks, doodles, etc. based on `rrrsa::` implementation 

```{r normalize}
normalize <- function(vec){
  # all elements of `vec` must be positive or zero 
  stopifnot(min(vec, na.rm=TRUE) >= 0)
  # if `vec` is all zeros, just return it right back 
  if (length(unique(vec))==1){
    if (unique(vec)==0){
      return(vec)
    }
  }
  # normalize by dividing each element by the sum 
  return(vec / sum(vec, na.rm=TRUE))
}
```


```{r normalize_matrix}
normalize_matrix <- function(mat, rows_or_cols){
  if (!rows_or_cols %in% c("rows", "cols")){
    message("gotta normalize over rows or cols mayne!")
    return(NULL)
  }
  if (rows_or_cols=="rows"){
    return(t(apply(mat, MARGIN=1, normalize)))
  } 
  if (rows_or_cols=="cols"){
    return(apply(mat, MARGIN=2, normalize))
  }
}
```



```{r utility}
utility <- function(items, costs, alpha){
  # `items` is a vector of nums, each representing a word/item 
  # `costs` must have same length as `items`, `alpha` a positive scalar 
  
  # apply `informativity()` to each item with specified costs and alpha. 
  # then collect them all and return with the same shape/order as `items`
  inform_unnorm <- mapply(informativity, prob=items, cost=costs, alpha=alpha)
  
  # normalization here is just dividing each number by the sum of them all 
  inform_norm <- normalize(inform_unnorm)
  
  return(inform_norm)
}

```


```{r informativity}
informativity <- function(prob, cost, alpha){
  # `prob` must be a probability, `alpha` a positive scalar 
  inform <- ifelse(prob == 0, 0, exp(-alpha * (-log(prob) - cost)))
  return(inform)
}

```


```{r full_recursion}
full_recursion <- function(mat, costs, priors, alpha){
  
  cost_mat <- matrix(
    nrow=nrow(mat), ncol=ncol(mat), 
    dimnames=dimnames(mat), 
    data=rep(as.numeric(costs), times=1, each=nrow(mat))
  )
  
  # the likelihood function is `utility()` 
  # utility is just normalized informativity 
  likelihood <- t(sapply(1:nrow(mat), function(row){
    utility(items=mat[row, ], costs=cost_mat[row, ], alpha=alpha)
  })) 
  
  unnorm_posterior <- apply(likelihood, MARGIN=2, function(col) priors * col)
  
  posterior <- normalize_matrix(unnorm_posterior, rows_or_cols="cols")
  
  dimnames(posterior) <- list(rownames(mat), colnames(mat))
  
  return(posterior)
}

```


```{r rsa_convenience}
rsa_matrix_to_df <- function(rsa_matrix, response_colname){
  out <- as.data.frame(rsa_matrix) %>% 
    mutate(quantity = as.numeric(gsub("row", "", rownames(rsa_matrix)))) %>% 
    melt(id.vars="quantity", variable.name="word") %>% 
    mutate(word = as.character(word)) %>% 
    arrange(quantity, word) 
    # rename(semantics = "value")
  
  names(out)[names(out)=="value"] <- response_colname
  
  return(out)
}


make_rsa_plot <- function(input_matrix, output_matrix, facet_words=TRUE){
  
  input_df <- rsa_matrix_to_df(input_matrix, response_colname="semantics")
  output_df <- rsa_matrix_to_df(output_matrix, response_colname="pragmatics")
  
  input_output <- 
    full_join(input_df, output_df, by=c("quantity", "word")) %>% 
    melt(id.vars=c("quantity", "word")) %>% 
    rename(type = variable) %>% 
    mutate(type = as.character(type))
  
  input_output$type <- factor(input_output$type, 
                              levels=c("semantics","pragmatics"))
  
  if (facet_words){
    the_plot <- 
      ggplot(input_output, aes(x=quantity, y=value, color=type)) + 
      geom_point(size=rel(.5)) + 
      geom_line() + 
      scale_color_manual(values=c("blue", "orange")) + 
      facet_wrap(~word)
    
  } else {
    the_plot <- 
      ggplot(input_output, aes(x=quantity, y=value, color=word)) + 
      geom_line() + 
      facet_wrap(~type)
  }
  
  return(the_plot)
}

rsaplot <- function(smat, costs, priors, alpha, facet_words=TRUE){
  if (!all(colSums(smat)==1)){
    smat <- normalize_matrix(smat, rows_or_cols="cols")
  }
  if (sum(priors)!=1){
    priors <- normalize(priors)
  }
  pmat <- full_recursion(smat, costs, priors, alpha)
  return(make_rsa_plot(smat, pmat, facet_words=facet_words))
}


```




```{r semantics_matrix}
words <- c("none","none_all","some","some_notall","most","all","not_all")
costs <- setNames(object=c(1, 1, 1, 1, 1, 1, 1), nm=words)

points <- 1:5
priors <- normalize(rep(1, length(points))) 

alpha <- 1

semantics_unnorm <- matrix(
  data=rep(NA, length(points)*length(words)), 
  nrow=length(points), ncol=length(words), 
  dimnames=list(paste0("row", points), words)
)
semantics_unnorm[, "none"]        <- c(1, 0, 0, 0, 0)
semantics_unnorm[, "none_all"]    <- c(1, 0, 0, 0, 1)
semantics_unnorm[, "some"]        <- c(0, 1, 1, 1, 1)
semantics_unnorm[, "some_notall"] <- c(0, 1, 1, 1, 0)
semantics_unnorm[, "most"]        <- c(0, 0, 1, 1, 1)
semantics_unnorm[, "all"]         <- c(0, 0, 0, 0, 1)
semantics_unnorm[, "not_all"]     <- c(1, 1, 1, 1, 0)
semantics_unnorm


rsaplot(semantics_unnorm, costs, priors, alpha)


s_a_na <- matrix(
  data=c(0,1,1,1,1,   0,0,0,0,1,    1,1,1,1,0), 
  nrow=5, ncol=3, dimnames=list(paste0("row", 1:5), 
                                c("some","all","not_all"))
)
rsaplot(s_a_na, costs=c(1,1,1), priors=rep(1, 5), alpha=1)

s_a <- matrix(
  data=c(0,1,1,1,1,   0,0,0,0,1), 
  nrow=5, ncol=2, dimnames=list(paste0("row", 1:5), 
                                c("some","all"))
)
rsaplot(s_a, costs=c(1,1), priors=rep(1, 5), alpha=1)

s_na <- matrix(
  data=c(0,1,1,1,1,   1,1,1,1,0), 
  nrow=5, ncol=2, dimnames=list(paste0("row", 1:5), 
                                c("some","not_all"))
)
rsaplot(s_na, costs=c(1,1), priors=rep(1, 5), alpha=1)

```



### A little calculus refresher

```{r}
# dnorm(x, mean=0, sd=1): height of unit normal density at x
plot(dnorm, from=-2, to=2)

points <- seq(from=-2, to=2, by=.1)

# proportion of "area under the curve" within the boundaries 
integrate(dnorm, lower=-1.96, upper=1.96)$value 
integrate(dnorm, -Inf, Inf)$value 
integrate(dnorm, -Inf, 0)$value; integrate(dnorm, 0, Inf)$value 

# a little calculus refresher 
dat <- data_frame(
  point=points, 
  norm_density=dnorm(points, mean=0, sd=1), 
  integral_nd_nd=sapply(norm_density, function(x){
    integrate(dnorm, lower=x-.05, upper=x+.05)$value
  }), 
  integral_nd_points=sapply(seq(from=-2, to=2, by=.1), function(x){
    integrate(dnorm, lower=x-.05, upper=x+.05)$value
  }), 
  deriv_points=as.numeric(eval(deriv(~ dnorm(points), "points")))
)
dat %>% 
  reshape2::melt(id.vars="point") %>% 
  ggplot(aes(x=point, y=value, color=variable)) + 
  geom_point() + facet_wrap(~variable, scales="free_y")

# watch out -- since sum of points is (effectively) zero! 
# sum(dat$point)
lapply(dat, function(x) x/sum(x)) %>% lapply(sum)

# plot(qnorm, from=-1, to=1)
# plot(pnorm, from=-2, to=1)


```



```{r eval=FALSE, include=FALSE}
### OLD CHUNK FROM PREVIOUS FIDDLE -- CONSOLIDATE AND/OR REMOVE AFTER REVIEWING

### try out all funcs on some basic vals
cost()                            # always maps to 1
e()                               # euler's num
eqn1(obj, utt, objects, words)    # .0952381
eqn2(utt, obj, objects, words)    # .6666667

s4_term(utt, objects)             # 1
satisfies(obj, utt, objects)      # TRUE

extension(utt, objects)           # "blue_circle"     
extension("square", objects)      # "blue_square", "green_square"

ll_interp(utt, obj, objects)     # TRUE
prior_obj(obj, objects)          # .3333333
prior_word(utt, words)           # .25

words_applying_to_obj(words, obj, objects)               # "blue", "circle"

prob_ref_given_word_in_context(obj, utt, objects, words) # .0952381
prob_word_given_ref_in_context(utt, obj, objects, words) # .6666667

sp_assertion(utt, obj, objects)  # word: "circle"; object: "blue_circle"; val=T

# GOTTA FINALIZE DIST FUNC IN `informativeness()`, which this calls
# *** utility(utt, obj, objects) ****************

# THIS ALSO NEEDS FIXED (RELATED TO `utility()`)
# *** informativeness(utt, dist_func=NA, obj)

# THIS NEEDS FIXED IN THE FONXE FILE BEFORE CAN BE USED...
# *** eqnS4(utt, obj, objects, words)   # 1.999856 <-- shd be same as eqn2()



```






<style>
<!-- /* use this to get font import code: http://fonts.googleapis.com/css?family=FONT+NAME*/ -->
<!-- /* use **latin**, not latin-extended */ -->
@font-face {
  font-family: 'Source Sans Pro';
  font-style: normal;
  font-weight: 400;
  src: local('Source Sans Pro Regular'), local('SourceSansPro-Regular'), url(https://fonts.gstatic.com/s/sourcesanspro/v11/ODelI1aHBYDBqgeIAH2zlJbPFduIYtoLzwST68uhz_Y.woff2) format('woff2');
  unicode-range: U+0000-00FF, U+0131, U+0152-0153, U+02C6, U+02DA, U+02DC, U+2000-206F, U+2074, U+20AC, U+2212, U+2215;
}


@font-face {
  font-family: 'Roboto Mono';
  font-style: normal;
  font-weight: 400;
  src: local('Roboto Mono'), local('RobotoMono-Regular'), url(https://fonts.gstatic.com/s/robotomono/v5/hMqPNLsu_dywMa4C_DEpY4gp9Q8gbYrhqGlRav_IXfk.woff2) format('woff2');
  unicode-range: U+0000-00FF, U+0131, U+0152-0153, U+02C6, U+02DA, U+02DC, U+2000-206F, U+2074, U+20AC, U+2212, U+2215;
}


body {
  padding: 10px;
  font-size: 12pt;
  font-family: 'Source Sans Pro', sans-serif;
}

h4 {
  margin-top: 32px;
  font-weight: bold;
}

blockquote {
  font-size: 12pt;
  font-family: 'Source Sans Pro', sans-serif;
}

code {
  font-family: 'Roboto Mono', monospace;
  font-size: 13px;
}

pre {
  font-family: 'Roboto Mono', monospace;
  font-size: 13px;
}

.table {
  width: 75%;
  align: center;
}


h3.subtitle {
  font-size: 26px;
  /*color: #aca39a;*/
  font-style: normal !important;
  font-weight: lighter !important;
  border-bottom: none;
  padding-top: 4px !important;
}

h4.author { 
  font-size: 22px;
  color: #aca39a;
  font-weight: lighter !important;
  padding-top: 0px !important;
  margin-top: 2px !important;
  margin-bottom: 12px !important;
  padding-bottom: 12px !important;
}

h3 {
  margin-top: 42px;
}

h2 {
  margin-top: 48px;
  border-bottom: 2px solid; 
}




</style>




